Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-05-10 paper.tex
Added a uses dep diagram
Added my name, fixed e-mails.
An initial stab at edits

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
	 \and 
	     Christoph Steinbeck$^6$%
	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
     \and 
         Ola Spjuth$^9$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
     \and 
         Jonathan Alvarsson$^9$%
         \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
     \and 
         Peter Murray-Rust$^7$%
         \email{Peter Murray-Rust - pm286@cam.ac.uk}
     \and 
         Daniel M Lowe$^7$%
         \email{Daniel Lowe - dl387@cam.ac.uk}
     \and
         Robert Hanson$^8$%
         \email{Robert Hanson - hansonr@stolaf.edu}
      \and
         Dmitry Pavlov$^8$%
         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
      \and
         Igor V Filippov\correspondingauthor$^{10}$%
         \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
    \iid(10)NIH, Somewhere in the US, USA
}%

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.

The formation of the Blue Obelisk group is somewhat unusual in that it
is not a funded network, nor does it follow the industry consortium
model. Rather it is a grassroots organisation, catalysed by an initial
core of interested scientists, but with membership open to all who
share one or more of the goals of the group:
\begin{itemize}
\item Open Source in Chemistry. One can use other people's code without further
permission, including changing it for one's own use and distributing
it again.
\item Open Standards in Chemistry. One can find visible community mechanisms for
protocols and communicating information. The mechanisms for creating
and maintaining these standards cover a wide spectrum of human
organisations, including various degrees of consent.
\item Open Data in Chemistry. One can obtain all data in the public domain when
wanted and reuse it for whatever purpose.
\end{itemize}

Note that while some may advocate also for Open Access to
publications, the goals focus more on the availability of code (to
reproduce results), standards (to exchange data), and the scientific
data itself. All three of these goals stem from the basics of the
scientific method for data sharing and reproducibilty.

The Blue Obelisk was first described in the CDK News \cite{CDKNewsBO} and
later as a formal paper by Guha et al.\cite{Guha2006} in
2006. This
contribution looks back on the past 5 years and surveys progress and
remaining challenges in the areas of Open Source, Open Data and Open
Standards in chemistry.

@@

Open Source toolkits for cheminformatics have now existed for nearly
ten years. During this period, some toolkits were developed from
scratch in academia, whereas others were made Open Source by releasing in-house
codebases under liberal licenses. When the Blue Obelisk was
established five years ago, the primary toolkits under active development
were the Chemistry Development Kit (CDK)
\cite{Steinbeck2003, Steinbeck2006}, Open Babel \cite{WebOpenBabel},
and JOELib \cite{WebJOELib}. Of these, both the CDK and Open Babel
continue to be actively developed. [Insert main focus of work on CDK
in last 5 years] Since 2006, major new features of Open Babel include 3D structure
generation and 2D structure-diagram generation, UFF and MMFF94
forcefields, and significantly expanded support for computational
chemistry calculations. In addition, a major focus of Open Babel development
has been to fix problems in the areas of stereochemistry, kekulisation
and canonicalisation.

Some project have also been adopted by other fields, including biology
and crystallography. A good example
here is Jmol, which has, in the past five years, gone through two major revisions
as part of an accelerated development focus. Starting with Jmol 10.2, released in
April of 2006, Jmol 11.0 added over 100 new features and was released in March of
2007. Three years later, Jmol 12.0 followed with over 700 new features.
Based on the outstanding work of Michael (Miguel) Howard and others, this development
path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
visualisation package, including full support for crystallography~\cite{Hanson2010},
efficient delivery of surface data using fast solvent-excluded surface
generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
 display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimisation using the UFF force field, and
a full implementation of Daylight SMILES and SMART, with extensions to
conformational and biomolecular substructure searching in the form of Jmol
BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
Behind the scenes, Jmol has been essentially completely rewritten. A full
mathematical expression evaluator was added, and the scripting language
was folded into a versatile JavaScript-like command language, including a
complete set of program flow commands, several new data types, and
user-defined variables and functions. An exciting non-chemical application
of Jmol has been its incorporation into the open source Sage
Mathematics Project [http://www.sagemath.org/], where Jmol is being used
to deliver three-dimensional depiction of mathematical functions.

@@

@@

@@

Another example of Open Source
tools originating from commercial groups is the ChemCraft tool from
Molecular Networks GmBH, that does XXX.

>>>>

\author{
  Jonathan Alvarsson$^1$%
  \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
  \and
  Igor V Filippov\correspondingauthor$^2$%
  \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
  \and
  Rajarshi Guha$^3$%
  \email{Rajarshi Guha - guhar@mail.nih.gov}
  \and
  Robert Hanson$^4$%
  \email{Robert Hanson - hansonr@stolaf.edu}
  \and
  Geoffrey R Hutchison$^5$%
  \email{Geoffrey R Hutchison - geoffh@pitt.edu}
  \and
  Daniel M Lowe$^6$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
  \and
  Peter Murray-Rust$^7$%
  \email{Peter Murray-Rust - pm286@cam.ac.uk}
  \and
  Noel M O'Boyle$^8$%
  \email{Noel M O'Boyle - n.oboyle@ucc.ie}
  \and
  Dmitry Pavlov$^9$%
  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
  \and
  Ola Spjuth$^1$%
  \email{ola.spjuth@farmbio.uu.se}
  \and
  Christoph Steinbeck$^{10}$%
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
  \and
  Egon L Willighagen$^{11}$%
  \email{Egon L Willighagen - egon.willighagen@ki.se}
    }

@@

\address{%
   \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA\\
    \iid(11)NIH, Somewhere in the US, USA
}%

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies
of small molecules was created as a resource about chemical structure and nomenclature
by biologists.\cite{DeMatos:2009p3839}

The formation of the Blue Obelisk group is somewhat unusual in that it
is not a funded network, nor does it follow the industry consortium
model. Rather it is a grassroots organisation, catalysed by an initial
core of interested scientists, but with membership open to all who
share one or more of the goals of the group:
\begin{itemize}
\item Open Data in Chemistry. One can obtain all scientific data in the public domain when
wanted and reuse it for whatever purpose.
\item Open Source in Chemistry. One can use other people's code without further
permission, including changing it for one's own use and distributing
it again.
\item Open Standards in Chemistry. One can find visible community mechanisms for
protocols and communicating information. The mechanisms for creating
and maintaining these standards cover a wide spectrum of human
organisations, including various degrees of consent.
\end{itemize}

Note that while some may advocate also for Open Access to
publications, the Blue Obelisk goals (ODOSOS) focus more on the availability of code (to
reproduce results), standards (to exchange data), and the scientific
data itself. All three of these goals stem from the fundamental
tenants of the scientific method for data sharing and reproducibilty.

The Blue Obelisk was first described in the CDK News \cite{CDKNewsBO} and
later as a formal paper by Guha et al.\cite{Guha2006} in
2006. This
contribution looks back on the past 5 years and surveys progress and
remaining challenges in the areas of Open Data, Open Source, and Open
Standards in chemistry.

@@

Open Source toolkits for cheminformatics have now existed for nearly
ten years. During this period, some toolkits were developed from
scratch in academia, whereas others were made Open Source by releasing in-house
codebases under liberal licenses. When the Blue Obelisk was
established five years ago, the primary toolkits under active development
were the Chemistry Development Kit (CDK)
\cite{Steinbeck2003, Steinbeck2006}, Open Babel \cite{WebOpenBabel},
and JOELib \cite{WebJOELib}. Of these, both the CDK and Open Babel
continue to be actively developed.

[Insert main focus of work on CDK in last 5 years]

Since 2006, major new features of Open Babel include 3D structure
generation and 2D structure-diagram generation, UFF and MMFF94
forcefields, and significantly expanded support for computational
chemistry calculations. In addition, a major focus of Open Babel development
has been to provide for accurate conversion and representation in
areas of stereochemistry, kekulisation, and canonicalisation. The
project has also grown, in terms of new contributors, new support from
commercial companies, and second-generation tools applying Open Babel
to a variety of end-user applications, from molecular editors to
chemical database systems.

@@

\subsubsection*{Web applications}

Some projects have also been adopted by other fields, including biology
and crystallography. A good example
here is Jmol, which has, in the past five years, gone through two major revisions
as part of an accelerated development focus. Starting with Jmol 10.2, released in
April of 2006, Jmol 11.0 added over 100 new features and was released in March of
2007. Three years later, Jmol 12.0 followed with over 700 new features.
Based on the outstanding work of Michael (Miguel) Howard and others, this development
path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
visualisation package, including full support for crystallography~\cite{Hanson2010},
efficient delivery of surface data using fast solvent-excluded surface
generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
 display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimisation using the UFF force field, and
a full implementation of Daylight SMILES[ref] and SMARTS[ref], with extensions to
conformational and biomolecular substructure searching in the form of Jmol
BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
Behind the scenes, Jmol has been essentially completely rewritten. A full
mathematical expression evaluator was added, and the scripting language
was folded into a versatile JavaScript-like command language, including a
complete set of program flow commands, several new data types, and
user-defined variables and functions. An exciting non-chemical application
of Jmol has been its incorporation into the open source Sage
Mathematics Project [http://www.sagemath.org/], where Jmol is being used
to deliver three-dimensional depiction of mathematical functions.

@@

% We should cite Craig James
% Care and Feeding of FIOSS: http://www.moonviewscientific.com/essays/software_lifecycle.htm
% And, of course, the Warren/Matt Stahl debate:
%Warren L. DeLano. "The case for open-source software in drug discovery" Drug Discovery Today (2005) 10(3) p. 213-217. DOI:10.1016/S1359-6446(04)03363-X
%Matthew T. Stahl. "Open-source software: not quite endsville" Drug Discovery Today (2005) 10(3) p. 219-222. DOI:10.1016/S1359-6446(04)03364-1

@@

Another example of Open Source
tools originating from commercial groups is the ChemCraft tool from
Molecular Networks GmBH, that does XXX.
% Really? I've only seen commercial software from them -GRH
